\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{array}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{xcolor}
\usepackage{listings}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  citecolor=blue
}

\definecolor{cacheirblue}{HTML}{1F4E79}
\definecolor{codebg}{HTML}{F5F7FA}
\definecolor{codeframe}{HTML}{D9E2EC}

\lstdefinestyle{cacheircode}{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{codeframe},
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false
}

\lstset{style=cacheircode}

\title{\textbf{CacheIR Project Documentation}\\
\large A Narrow, Explainable Compiler and Runtime for Transformer Inference}
\author{CacheIR Contributors}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
CacheIR is a focused machine-learning compiler and runtime for decoder-only
transformer inference. It imports Llama, Mistral, and Qwen-style model metadata,
lowers the model into a custom intermediate representation, applies
transformer-specific optimization passes, specializes prefill and decode graphs,
plans memory and execution schedules, selects backend kernels, and executes
inference through an independent reference runtime. The project is intentionally
not a wrapper around PyTorch and not a clone of vLLM. Its primary design goal is
to make the complete path from model graph to runtime schedule inspectable,
testable, and extensible.
\end{abstract}

\tableofcontents
\newpage

\section{Project Purpose}

General ML compiler frameworks such as TVM, MLIR-based stacks, IREE, and
StableHLO-oriented pipelines are broad systems designed to support many models
and many hardware targets. Large language model serving has a narrower and very
distinct execution profile. Prefill processes an entire prompt and tends to be
compute heavy. Decode generates one token at a time and tends to be dominated by
KV-cache traffic, memory bandwidth, batching policy, and attention kernel choice.

CacheIR exists to explore that narrower problem directly. The project focuses on
five ideas:

\begin{enumerate}[leftmargin=*]
  \item Treat prefill and decode as separate compiler modes.
  \item Compile transformer operations and KV-cache policy together.
  \item Expose each optimization pass as a textual graph diff.
  \item Emit memory plans, hardware hints, and execution schedules as first-class
        artifact data.
  \item Provide a real CPU reference runtime path that does not execute through
        PyTorch, plus an optional CUDA backend where CacheIR still owns graph
        execution and PyTorch supplies the tensor/cuBLAS substrate.
\end{enumerate}

The current implementation is a complete reference project. It includes an
installable Python package, command-line interface, model importer skeletons,
CacheIR graph IR, compiler passes, artifact bundles, graph visualization,
benchmarks, a NumPy CPU runtime, an end-to-end CUDA artifact runtime, a paged
KV-cache reference implementation, an OpenAI-compatible local server with health
and metrics endpoints, prefix-cache reuse with counters, a small continuous-batch
scheduler with CUDA batched prefill/decode paths, calibrated spillover
experiments, native GGUF dense and classic
quantized block reads, StableHLO and MLIR-style round-trip experiments, a
C++20/OpenMP AVX backend with an optional pybind11 bridge, guarded Triton
kernels, optional CUDA fused-kernel sources, FP16 Tensor Core matmul experiments,
IREE/TVM upstream benchmark bridges, accelerator adapter probes, Docker support,
CI, and tests.

\section{High-Level Architecture}

CacheIR is organized into five layers:

\begin{longtable}{p{0.18\linewidth}p{0.76\linewidth}}
\toprule
\textbf{Layer} & \textbf{Role} \\
\midrule
Model importer & Reads model configuration, tensor metadata, tokenizer metadata,
rope settings, attention layout, shape information, and quantization hints. \\
Intermediate representation & Represents the model as CacheIR graph nodes with
explicit tensor types, weights, attributes, graph modes, layouts, schedules, and
memory plans. \\
Optimization passes & Transform the logical graph into optimized, mode-specific
graphs for prefill and decode. \\
Kernel backend & Selects CPU or GPU-oriented kernels and records hardware-aware
execution hints. \\
Runtime & Loads compiled artifacts, weights, tokenizers, and KV-cache state, then
executes scheduled graph nodes through backend dispatch. \\
\bottomrule
\end{longtable}

\subsection{Repository Layout}

\begin{lstlisting}
cacheir/
  ir.py                  CacheIR graph IR
  compiler.py            importer selection and pass pipeline
  importers/             HF, ONNX, GGUF, StableHLO, tiny model support
  passes/                compiler optimization passes
  runtime/               artifacts, tokenizer, KV cache, CPU/CUDA runtimes, scheduler, server
  backends/              registry, native bridge, Triton kernels, upstream probes
  benchmark.py           benchmark runner
  visualize.py           HTML/DOT/IR/MLIR graph export
  hardware.py            hardware profile discovery
  quantization.py        quantization simulation utilities
cpp/                     C++20/OpenMP AVX backend and pybind11 module
docs/                    Markdown and LaTeX documentation
examples/                tiny model compile demo
scripts/                 benchmark matrix, scheduler, GPU, and external comparison utilities
tests/                   pytest coverage
\end{lstlisting}

\section{Model Import}

\subsection{Supported Import Paths}

CacheIR currently supports several import paths at different maturity levels:

\begin{longtable}{p{0.24\linewidth}p{0.18\linewidth}p{0.50\linewidth}}
\toprule
\textbf{Format} & \textbf{Status} & \textbf{Details} \\
\midrule
Hugging Face directory & Reference path & Reads \texttt{config.json}, discovers
NPZ or safetensors tensor shapes, and builds a Llama/Mistral/Qwen-style decoder
graph. \\
NPZ tiny model & Reference test path & Used by \texttt{cacheir make-tiny}, tests,
examples, and benchmarks. \\
safetensors & Metadata path & Shape discovery and runtime tensor loading
are supported when the optional \texttt{safetensors} dependency is installed. \\
ONNX & Generic graph skeleton & Imports generic ONNX nodes into CacheIR ops where
straightforward. Full transformer semantic recovery is not claimed. \\
GGUF & Metadata plus native subset & Parses GGUF header metadata and tensor
directory, maps common Llama tensor names, builds a CacheIR graph, and can read
dense \texttt{F32}/\texttt{F16}/\texttt{BF16}/\texttt{I8}/\texttt{I16}/\texttt{I32}/\texttt{I64}/\texttt{F64} plus \texttt{Q4\_0}/\texttt{Q4\_1}/\texttt{Q5\_0}/\texttt{Q5\_1}/\texttt{Q8\_0}/\texttt{Q8\_1}
tensor payloads directly for native execution experiments. K/IQ/TQ/NV/MX
families are delegated to the optional upstream \texttt{gguf} reference package
when it exposes a dequantizer. \\
StableHLO text & Experimental subset & Parses textual StableHLO arithmetic,
shape, broadcast, slice, reduce, and cast operations and maps them into CacheIR
nodes for importer experiments. \\
\bottomrule
\end{longtable}

\subsection{Model Configuration}

The central importer data structure is \texttt{ModelConfig}. It records:

\begin{itemize}[leftmargin=*]
  \item architecture name
  \item hidden size
  \item intermediate MLP size
  \item number of layers
  \item number of attention heads
  \item number of key/value heads
  \item vocabulary size
  \item maximum position embeddings
  \item RoPE theta
  \item RMSNorm epsilon
  \item dtype
  \item tied embedding policy
  \item quantization hints
\end{itemize}

This configuration is used to construct a logical decoder graph. For each layer,
the importer emits attention norm, Q/K/V projections, RoPE, grouped-query
attention, output projection, residual adds, MLP norm, gate/up projections, SiLU,
elementwise multiplication, down projection, and final normalization/head nodes.

\section{Intermediate Representation}

\subsection{Design Goals}

CacheIR's IR is intentionally small and text-friendly. It is not trying to be a
general replacement for MLIR, StableHLO, or TVM Relay. Instead, it represents the
subset of operations needed to compile and run decoder-only transformer inference
while preserving enough metadata for scheduling, memory planning, and pass-level
explainability.

\subsection{Core Data Types}

\begin{longtable}{p{0.24\linewidth}p{0.68\linewidth}}
\toprule
\textbf{Type} & \textbf{Purpose} \\
\midrule
\texttt{TensorType} & Shape, dtype, layout, byte-size estimation, and symbolic
dimension handling. \\
\texttt{WeightSpec} & Maps an IR value to a named tensor key, tensor type,
optional quantization metadata, and optional backing file. \\
\texttt{Node} & Operation, inputs, outputs, attributes, name, and side-effect
flag. \\
\texttt{Graph} & Complete IR graph with inputs, outputs, nodes, values, weights,
constants, graph attributes, mode, and target. \\
\texttt{CompileArtifact} & Serializable package containing graphs, pass traces,
model config, target, quantization, and metadata. \\
\bottomrule
\end{longtable}

\subsection{Important IR Operations}

\begin{longtable}{p{0.30\linewidth}p{0.62\linewidth}}
\toprule
\textbf{Operation} & \textbf{Meaning} \\
\midrule
\texttt{token\_embedding} & Gathers hidden vectors from token IDs and embedding
weights. \\
\texttt{rms\_norm} & RMS normalization with learned scale. \\
\texttt{matmul} & Linear projection using out-by-in weight layout. \\
\texttt{qkv\_projection} & Fused logical Q/K/V projection produced by compiler
passes. \\
\texttt{rope} & Rotary position embedding applied to Q and K. \\
\texttt{grouped\_query\_attention} & Logical GQA/MQA attention before mode
specialization. \\
\texttt{paged\_attention\_prefill} & Prefill-specialized attention op. \\
\texttt{paged\_attention\_decode} & Decode-specialized attention op. \\
\texttt{fused\_rmsnorm\_qkv\_rope} & Fused norm, Q/K/V projection, and RoPE op. \\
\texttt{fused\_swiglu} & Fused gate projection, up projection, SiLU, and multiply. \\
\texttt{quantized\_matmul} & Quantization-aware matmul lowering. \\
\bottomrule
\end{longtable}

\subsection{Example IR}

Logical attention begins as separate nodes:

\begin{lstlisting}
%attn_norm = rms_norm(%hidden, %attn_norm_w)
%q = matmul(%attn_norm, %q_proj_w)
%k = matmul(%attn_norm, %k_proj_w)
%v = matmul(%attn_norm, %v_proj_w)
%q_rope, %k_rope = rope(%q, %k)
%attn = grouped_query_attention(%q_rope, %k_rope, %v)
\end{lstlisting}

After fusion and decode specialization, the graph contains fewer scheduled calls:

\begin{lstlisting}
%q_rope, %k_rope, %v = fused_rmsnorm_qkv_rope(...)
%attn = paged_attention_decode(%q_rope, %k_rope, %v)
\end{lstlisting}

\section{Compiler Pipeline}

\subsection{Pass Manager}

The pass manager applies a fixed transformer-oriented pipeline to each requested
execution mode. A compile call can request both \texttt{prefill} and
\texttt{decode}; each mode receives a cloned graph and its own pass trace.

\subsection{Passes}

\begin{longtable}{p{0.28\linewidth}p{0.64\linewidth}}
\toprule
\textbf{Pass} & \textbf{Effect} \\
\midrule
Shape inference & Infers tensor types for values produced by each op. \\
Constant folding & Evaluates small constant arithmetic and activation nodes. \\
QKV projection fusion & Replaces separate Q, K, and V matmuls with one logical
\texttt{qkv\_projection}. \\
RMSNorm/QKV/RoPE fusion & Fuses RMSNorm, QKV projection, and RoPE into
\texttt{fused\_rmsnorm\_qkv\_rope}. \\
SwiGLU fusion & Fuses gate/up projection, SiLU, and elementwise multiply into
\texttt{fused\_swiglu}. \\
Prefill/decode specialization & Replaces logical attention with
\texttt{paged\_attention\_prefill} or \texttt{paged\_attention\_decode}. \\
Layout conversion & Selects CPU or GPU-oriented activation and weight layouts. \\
Quantization-aware lowering & Rewrites matmul-family ops to quantized variants
when a quantization mode is requested. \\
Dead node elimination & Removes values that no longer contribute to graph
outputs. \\
Hardware-adaptive planning & Records page sizes, low-VRAM hints, and cost-class
metadata based on local hardware profile and target. \\
Kernel selection & Assigns backend kernel names to graph nodes. \\
Execution planning & Emits a schedule with step number, op, kernel, inputs,
outputs, estimated bytes, and estimated FLOPs. \\
Static memory planning & Computes a reusable activation arena with offsets,
sizes, and value lifetimes. \\
\bottomrule
\end{longtable}

\subsection{Pass Trace and Explainability}

Every pass records:

\begin{itemize}[leftmargin=*]
  \item pass name
  \item whether the graph changed
  \item remarks
  \item full before text
  \item full after text
  \item unified diff
\end{itemize}

This makes the compiler inspectable from the command line:

\begin{lstlisting}
cacheir inspect examples/tiny_artifact \
  --mode decode \
  --pass-name prefill_decode_specialization
\end{lstlisting}

\section{Prefill and Decode Specialization}

Prefill and decode are not treated as one static graph. CacheIR compiles separate
graphs because the phases stress different resources.

\subsection{Prefill}

Prefill consumes the prompt sequence. It uses causal masking across the prompt and
is relatively compute heavy. The prefill graph records
\texttt{paged\_attention\_prefill} nodes and a memory plan sized for prompt-length
activations.

\subsection{Decode}

Decode usually consumes one token per step. It appends keys and values to the
KV-cache and repeatedly attends over the cache. The decode graph records
\texttt{paged\_attention\_decode} nodes, smaller activation memory requirements,
decode-oriented page-size hints, and cost-class labels such as
\texttt{kv\_bandwidth\_bound}.

\section{KV-Cache-Aware Runtime}

\subsection{Paged KV Cache}

The runtime includes a \texttt{PagedKVCache}. The reference CPU implementation
stores keys and values in contiguous NumPy arrays for simplicity, but it also
maintains page metadata:

\begin{itemize}[leftmargin=*]
  \item page size
  \item page IDs
  \item layer-to-page mapping
  \item token offsets and page lengths
\end{itemize}

This gives the compiler and runtime a stable policy contract that Triton and CUDA
kernels can consume.

\subsection{Prefix Cache and Continuous Batching}

The runtime includes a small LRU \texttt{PrefixCache}. It stores KV-cache
snapshots keyed by token prefixes and records hit, miss, and eviction counters.
Prefill can restore a cached prefix and append only the uncached suffix while
preserving RoPE offsets and causal masking. This makes prefix reuse observable
and testable rather than a hidden serving optimization.

The \texttt{ContinuousBatchScheduler} creates independent per-request KV-cache
sessions with \texttt{Runtime.fork()}, while sharing weights, tokenizer state,
and the prefix cache. The scheduler advances active requests in decode rounds
and records submitted requests, completed requests, generated tokens,
prefill/decode milliseconds, reused prefix tokens, scheduling rounds, and maximum
observed batch size. The reference CPU backend still executes each request
independently, but the scheduling contract mirrors the control surface a future
shared GPU paged-KV runtime would consume.

For CUDA sessions, the scheduler can batch variable-length prompt prefill when
prefix reuse is disabled. Dense graph work runs over a padded batch, while
attention and KV writes are sliced back to each request's real sequence length.
Active decode rounds advance through \texttt{run\_decode\_batch}. The current
implementation batches norm, matmul, residual, and MLP work while preserving
each request's independent K/V tensors at attention nodes. Forked CUDA sessions
share page allocator accounting, so page IDs and resident-page statistics live
in one runtime namespace. The scheduler also supports queue limits, request
priorities, queued cancellation, and admission-control metrics. This is not yet
the final production design with persistent page-backed shared KV storage, but
it is a real scheduler-level CUDA batching path.

\subsection{Calibrated Spillover Cost Model}

The spillover policy can be driven by a \texttt{SpilloverCostModel}. It records
page bytes, free GPU memory, safety margin, PCIe bandwidth, CPU/GPU bandwidth
estimates, and transfer latencies. From those inputs it computes a resident-page
budget and annotates spilled pages with target and estimated transfer cost. This
keeps low-VRAM experiments deterministic in tests while still allowing hardware
profiles or measured bandwidths to tighten the policy on real machines.

\subsection{Runtime Flow}

The runtime performs the following steps:

\begin{enumerate}[leftmargin=*]
  \item Load a \texttt{CompileArtifact}.
  \item Create a \texttt{WeightStore}.
  \item Load tokenizer metadata through \texttt{TokenizerBridge}.
  \item Allocate a \texttt{PagedKVCache}.
  \item Convert input text to token IDs.
  \item Execute the prefill graph.
  \item Repeatedly execute the decode graph for greedy generation.
  \item Optionally schedule multiple requests through
        \texttt{ContinuousBatchScheduler}.
\end{enumerate}

\section{Backend Model}

\subsection{CPU Reference Backend}

The CPU backend is a NumPy reference runtime. It implements:

\begin{itemize}[leftmargin=*]
  \item token embedding
  \item RMSNorm
  \item matmul
  \item fused RMSNorm/QKV/RoPE
  \item RoPE
  \item paged attention prefill and decode
  \item residual add
  \item SiLU
  \item elementwise multiply
  \item fused SwiGLU
  \item softmax
\end{itemize}

This backend is designed for correctness, compiler validation, artifact testing,
and benchmark regression tracking. It is not presented as a high-performance LLM
serving backend.

\subsection{C++ Backend and pybind11 Bridge}

The \texttt{cpp/} directory contains a C++20/OpenMP backend with kernels for
RMSNorm, SiLU, fused SiLU-multiply, and out-by-in matmul. The matmul and RMSNorm
paths use a runtime dot-product dispatcher with scalar, AVX2/FMA, and AVX512
implementations where the compiler and host CPU support them. The static library
builds with CMake and Ninja:

\begin{lstlisting}
cmake -S cpp -B cpp/build -DCMAKE_BUILD_TYPE=Release
cmake --build cpp/build --config Release
\end{lstlisting}

The optional pybind11 module exposes native RMSNorm, fused SiLU-multiply,
out-by-in matmul, and SIMD capability reporting to Python:

\begin{lstlisting}
python -m pip install -e ".[native]"
cmake -S cpp -B cpp/build -DCACHEIR_BUILD_PYTHON=ON -DCMAKE_BUILD_TYPE=Release
cmake --build cpp/build --config Release
\end{lstlisting}

On Windows, pinning \texttt{Python3\_EXECUTABLE} to the active interpreter avoids
CMake choosing another Python installation. The latest local run installed
\texttt{pybind11} from PyPI and built \texttt{\_cacheir\_native} successfully.
The Python runtime dispatches native RMSNorm and fused SiLU-multiply
automatically when available. Native matmul remains opt-in through
\texttt{CACHEIR\_NATIVE\_MATMUL=force} or \texttt{CACHEIR\_NATIVE\_MATMUL=auto}
because NumPy's optimized contraction is faster for many local decode shapes.

\subsection{CUDA Artifact Runtime}

\texttt{CudaRuntime} executes the same compiled CacheIR artifacts on CUDA
tensors. It is not a replacement compiler pipeline and it does not hand execution
to an upstream serving engine. CacheIR still owns graph traversal, KV-cache
state, tokenizer integration, and kernel dispatch. PyTorch CUDA is used as the
GPU tensor substrate and cuBLAS access path, while CacheIR dispatches its own
Triton elementwise kernels where useful.

The CUDA runtime includes:

\begin{itemize}[leftmargin=*]
  \item lazy CPU-to-GPU weight loading with fp16 or fp32 runtime dtype selection
  \item GPU-resident KV-cache state with the same page metadata contract as the
        CPU reference cache
  \item artifact-driven prefill and decode graph execution
  \item SDPA attention dispatch, which can use optimized CUDA attention kernels
        available in the installed PyTorch build
  \item guarded Triton RMSNorm and SiLU-multiply dispatch for larger tensors
  \item cached packed QKV and Gate/Up weights, reducing decode matmul launches
        for fused CacheIR nodes
  \item \texttt{cacheir benchmark --backend cuda --cuda-dtype float16} support
\end{itemize}

This backend moves CacheIR beyond isolated GPU kernel microbenchmarks: complete
decoder artifacts can now execute on the RTX 5070 Laptop GPU. It is still not a
production serving backend because persistent page-backed KV storage used
directly by fused attention kernels, packed quantized GEMM, CUDA graph replay,
and large-model coverage remain future work.

\subsection{GPU and Triton Target Surface}

The compiler can target \texttt{cuda} or \texttt{triton} for kernel naming and
schedule generation. The project includes guarded Triton kernels for RMSNorm and
SiLU/SwiGLU activation multiply, fused RMSNorm/QKV/RoPE projection,
single-query decode attention, and multi-batch page-table decode attention with
GQA mapping when the optional \texttt{triton} package and a CUDA-capable
environment are available. It also includes a guarded Triton FP16 matmul kernel
that lowers through \texttt{tl.dot}. The \texttt{cpp/} tree includes CUDA C++
fused kernels for RMSNorm, SwiGLU, RMSNorm/QKV/RoPE, FP16 WMMA Tensor Core
matmul, a batched paged-attention ABI, a reduced block-level paged-attention
decode path, and CUDA graph capture planning.

\subsection{Accelerator Adapter Probes}

CacheIR includes guarded adapter probes and direct wrappers for CUTLASS,
FlashAttention, and FlashInfer. These adapters do not replace the compiler pipeline: CacheIR still
owns IR, memory planning, schedules, and dispatch decisions. When an optional
package is importable, a compatible scheduled op can be routed to that external
library for experiments and benchmark comparison. CacheIR exposes guarded direct
wrappers for FlashAttention prefill, FlashInfer single decode, and FlashInfer
batch paged decode wrapper classes. On the current Windows verification machine,
\texttt{nvidia-cutlass 4.2.0.0} installs and is detected through
\texttt{cutlass\_cppgen}. Native Windows/Python 3.13 still lacks a complete
wheel set for vLLM, FlashAttention, FlashInfer, and llama-cpp-python, so the
CUDA validation path uses WSL2 environments. Those environments executed real
FlashAttention prefill and FlashInfer decode smoke paths through CacheIR's
adapter wrappers.

\section{Quantization}

CacheIR supports quantization-aware lowering. When a compile call requests a
quantization mode such as \texttt{int4\_awq}, matmul-family ops are rewritten to
quantized variants:

\begin{itemize}[leftmargin=*]
  \item \texttt{matmul} becomes \texttt{quantized\_matmul}
  \item \texttt{fused\_rmsnorm\_qkv\_rope} becomes
        \texttt{quantized\_fused\_rmsnorm\_qkv\_rope}
  \item \texttt{fused\_swiglu} becomes \texttt{quantized\_fused\_swiglu}
\end{itemize}

The CPU reference runtime simulates quantized weights by quantizing and
dequantizing weights once on load. This validates graph lowering and runtime
compatibility but is not a compressed high-performance kernel implementation.

\section{Artifacts}

\subsection{Single-File Artifact}

A \texttt{CompileArtifact} can be saved as JSON. It includes:

\begin{itemize}[leftmargin=*]
  \item artifact version and creation time
  \item target
  \item quantization mode
  \item model path
  \item model configuration
  \item graphs by mode
  \item pass traces by mode
  \item compiler metadata
\end{itemize}

\subsection{Bundle Artifact}

When the output path has no file suffix, CacheIR writes a bundle:

\begin{lstlisting}
artifact.json
manifest.json
README.txt
graphs/prefill.cir
graphs/decode.cir
schedules/prefill.json
schedules/decode.json
passes/*.diff
\end{lstlisting}

This makes the compiler output reviewable without writing custom tooling.

\section{Command-Line Interface}

CacheIR exposes the following commands:

\begin{longtable}{p{0.20\linewidth}p{0.70\linewidth}}
\toprule
\textbf{Command} & \textbf{Purpose} \\
\midrule
\texttt{profile} & Print local CPU/GPU hardware profile. \\
\texttt{make-tiny} & Generate a small Llama-shaped test model. \\
\texttt{compile} & Compile a model into a CacheIR artifact. \\
\texttt{inspect} & Print optimized IR or a specific pass diff. \\
\texttt{export} & Export IR as HTML, DOT, or text. \\
\texttt{benchmark} & Measure prefill and decode performance separately. \\
\texttt{external} & Probe optional upstream systems and run installed IREE/TVM
smoke benchmarks. \\
\texttt{run} & Run greedy local generation. \\
\texttt{serve} & Start an OpenAI-compatible local server. \\
\bottomrule
\end{longtable}

\subsection{Example Session}

\begin{lstlisting}
cacheir make-tiny examples/tiny_model
cacheir compile examples/tiny_model --output examples/tiny_artifact
cacheir inspect examples/tiny_artifact --mode decode
cacheir benchmark examples/tiny_artifact --decode-tokens 32 --repeats 3
cacheir benchmark examples/tiny_artifact --backend cuda --cuda-dtype float16 --warmup 2
cacheir external --benchmark --workdir .tmp/upstream
cacheir export examples/tiny_artifact examples/decode.html --mode decode
cacheir run examples/tiny_artifact --prompt "CacheIR"
\end{lstlisting}

\section{Python API}

The public API is intentionally small:

\begin{lstlisting}[language=Python]
from cacheir import Runtime, compile_model

artifact = compile_model(
    model_path="examples/tiny_model",
    target="cpu",
    quant=None,
    mode=["prefill", "decode"],
    max_seq=128,
)

rt = Runtime(artifact)
for token in rt.generate("Explain MLIR in simple terms", max_new_tokens=16):
    print(token, end="")
\end{lstlisting}

\section{Serving API}

When optional server dependencies are installed, CacheIR can start a FastAPI
server with OpenAI-compatible endpoints:

\begin{lstlisting}
cacheir serve examples/tiny_artifact --host 127.0.0.1 --port 8000
\end{lstlisting}

Supported endpoints include:

\begin{itemize}[leftmargin=*]
  \item \texttt{GET /v1/models}
  \item \texttt{POST /v1/completions}
  \item \texttt{POST /v1/chat/completions}
\end{itemize}

Streaming chat completions return server-sent events with OpenAI-style JSON
chunks.

\section{Benchmarking}

\subsection{Single Artifact Benchmark}

\begin{lstlisting}
cacheir benchmark examples/tiny_artifact \
  --prompt "CacheIR benchmark" \
  --decode-tokens 64 \
  --repeats 5
\end{lstlisting}

The output reports:

\begin{itemize}[leftmargin=*]
  \item prompt token count
  \item decode token count
  \item average prefill latency
  \item average decode latency per token
  \item prefill tokens per second
  \item decode tokens per second
  \item KV-cache page statistics
\end{itemize}

\subsection{CUDA Runtime Benchmark}

The dedicated CUDA runtime benchmark compares the CPU reference backend and the
CUDA artifact backend through the same benchmark API:

\begin{lstlisting}
python scripts/benchmark_cuda_runtime.py \
  --output .tmp/reports/cuda_runtime_benchmark_h1024_latest.json \
  --repeats 3 \
  --warmup 1 \
  --decode-tokens 8 \
  --hidden-size 1024 \
  --num-layers 4 \
  --cuda-dtype float16
\end{lstlisting}

\begin{longtable}{p{0.22\linewidth}p{0.18\linewidth}p{0.14\linewidth}p{0.16\linewidth}p{0.14\linewidth}p{0.14\linewidth}}
\toprule
\textbf{Shape} & \textbf{Backend} & \textbf{Prefill ms} & \textbf{Decode ms/tok} &
\textbf{Prefill tok/s} & \textbf{Decode tok/s} \\
\midrule
h128, 2 layers & CPU fp32 & 2.122 & 1.007 & 17,432.8 & 993.0 \\
h128, 2 layers & CUDA fp16 & 3.058 & 3.210 & 12,099.6 & 311.5 \\
h512, 2 layers & CPU fp32 & 5.828 & 2.077 & 6,348.3 & 481.4 \\
h512, 2 layers & CUDA fp16 & 3.867 & 4.030 & 9,567.5 & 248.1 \\
h1024, 4 layers & CPU fp32 & 30.984 & 12.271 & 1,194.1 & 81.5 \\
h1024, 4 layers & CUDA fp16 & 7.533 & 7.573 & 4,912.0 & 132.1 \\
h2048, 4 layers & CPU fp32 & 94.036 & 30.619 & 393.5 & 32.7 \\
h2048, 4 layers & CUDA fp16 & 6.705 & 7.333 & 5,517.9 & 136.4 \\
\bottomrule
\end{longtable}

The CUDA backend is slower on very small toy shapes where launch overhead
dominates. It crosses over as the model becomes compute-heavy: h1024/l4 is
4.11x faster on prefill and 1.62x faster on decode, while h2048/l4 is 14.02x
faster on prefill and 4.18x faster on decode. The result proves end-to-end GPU
model execution, not production parity with mature LLM serving engines.

\subsection{Benchmark Matrix}

The repository includes \texttt{scripts/benchmark\_matrix.py}, which generates
three reference model sizes, compiles fp32 and simulated int4 variants, runs three
prompt lengths, and emits JSON plus Markdown summaries:

\begin{lstlisting}
python scripts/benchmark_matrix.py \
  --output benchmark_results.json \
  --repeats 3 \
  --decode-tokens 16
\end{lstlisting}

\subsection{Scheduler Benchmark}

The repository also includes \texttt{scripts/benchmark\_scheduler.py}. It
creates or loads a CacheIR artifact, submits prompts to
\texttt{ContinuousBatchScheduler}, and reports request throughput, generated
token throughput, prefix reuse, and scheduling metrics:

\begin{lstlisting}
python scripts/benchmark_scheduler.py \
  --output .tmp/reports/scheduler_benchmark.json \
  --repeats 3 \
  --max-new-tokens 8 \
  --max-batch-size 4
\end{lstlisting}

\subsection{CUDA Scheduler Benchmark}

The CUDA scheduler benchmark compares sequential CUDA sessions with
variable-length batched prefill and active-request batched decode:

\begin{lstlisting}
python scripts/benchmark_cuda_scheduler.py \
  --output .tmp/reports/cuda_scheduler_benchmark_latest.json \
  --repeats 3 \
  --warmup 1 \
  --max-batch-size 4 \
  --max-new-tokens 16 \
  --hidden-size 1024 \
  --num-layers 4 \
  --cuda-dtype float16
\end{lstlisting}

\begin{longtable}{p{0.24\linewidth}p{0.26\linewidth}p{0.18\linewidth}p{0.18\linewidth}}
\toprule
\textbf{Shape} & \textbf{Mode} & \textbf{Median ms} & \textbf{Gen tok/s} \\
\midrule
h1024/l4, batch 4 & Sequential & 582.177 & 109.9 \\
h1024/l4, batch 4 & Batched prefill + decode & 339.796 & 188.3 \\
h2048/l4, batch 4 & Sequential & 867.240 & 37.0 \\
h2048/l4, batch 4 & Batched prefill + decode & 684.002 & 46.8 \\
\bottomrule
\end{longtable}

The h1024/l4 run improves latency and generated-token throughput by 1.71x. The
h2048/l4 run improves by 1.27x. These runs processed 134 real prefill tokens and
6 padding tokens. The shorter h2048 decode window leaves prefill as the dominant
cost, and attention still splits per request after the dense padded graph work.

\subsection{Installed Upstream Benchmarks}

When the optional upstream packages are installed, CacheIR can compile and run
small real upstream benchmark probes without user-supplied commands:

\begin{lstlisting}
cacheir external --benchmark --workdir .tmp/upstream

python scripts/compare_external_benchmarks.py examples/tiny_artifact \
  --run-installed-smoke \
  --output benchmark_comparison.json
\end{lstlisting}

The IREE path compiles StableHLO text to a VM flatbuffer through
\texttt{iree-base-compiler} and runs \texttt{iree-benchmark-module}. The TVM
path builds and runs a TE/TIR vector-add module through the installed
\texttt{apache-tvm} wheel. When model artifacts are available, CacheIR can also
launch model-aware vLLM and llama.cpp helpers:

\begin{lstlisting}
cacheir external --benchmark --workdir .tmp/upstream \
  --vllm-model /path/to/hf_model \
  --llama-model /path/to/model.gguf
\end{lstlisting}

These helpers run the installed upstream tools rather than wrapping them inside
CacheIR's runtime. The vLLM helper drives \texttt{vllm bench latency}; the
llama.cpp helper drives \texttt{llama-bench}.

\subsection{Local Native Experiment Snapshot}

The latest local verification on 2026-06-25 installed the full optional project
stack with \texttt{python -m pip install -e ".[dev,server,importers,benchmark,native,gpu]"}.
The native pybind11 module built successfully for Python 3.13 and reported
\texttt{simd\_backend() == "avx512"}. Native RMSNorm, matmul, and fused
SiLU-multiply matched NumPy within float32 tolerance. The runtime now dispatches
native RMSNorm and fused SiLU-multiply by default, while native matmul is
opt-in because NumPy's optimized contraction is faster for many local decode
shapes.

The CPU benchmark matrix ran 18 rows covering three tiny model sizes, fp32 and
simulated int4 variants, three prompt lengths, three repeats, and 16 decode
tokens. Selected results: \texttt{tiny\_1l\_h16} fp32 short decode was
0.571 ms/token, \texttt{small\_2l\_h32} fp32 short decode was 0.974 ms/token,
and \texttt{medium\_4l\_h64} fp32 medium decode was 2.169 ms/token.

The scheduler benchmark processed four requests and 32 generated tokens in a
median 54.819 ms, or 72.97 requests/s and 583.74 generated tokens/s, while
reusing 70 prompt tokens through the prefix cache.

The CUDA scheduler benchmark executed four CUDA sessions with variable-length
batched prefill and active-request batched decode. On h1024/l4, batch-4 latency
improved from 582.177 ms to 339.796 ms and generated-token throughput improved
from 109.9 tok/s to 188.3 tok/s. On h2048/l4, batch-4 latency improved from
867.240 ms to 684.002 ms and throughput improved from 37.0 tok/s to
46.8 tok/s. A smaller h128/l1 Triton batched-decode smoke run also executed
through the scheduler path with \texttt{--use-triton-attention}.

CUDA Torch \texttt{2.12.1+cu130} was available on the RTX 5070 Laptop GPU. The
Triton kernel benchmark measured RMSNorm at 0.0157 ms, SiLU-multiply at
0.0213 ms, FP16 matmul at 0.0667 ms or 8.05 TFLOP/s, single-query decode
attention at 0.0552 ms, and multi-batch page-table decode attention at
0.0566 ms. Torch matmul remained faster at 0.0275 ms or 19.54 TFLOP/s for the
same FP16 GEMM shape.

The end-to-end CUDA runtime benchmark executed full decoder artifacts with fp16
weights, GPU-resident KV state, SDPA attention, Triton elementwise dispatch
gates, and cached packed QKV/Gate-Up weights. The h1024/l4 CUDA fp16 run
measured 7.533 ms prefill and 7.573 ms/token decode versus CPU fp32 at
30.984 ms prefill and 12.271 ms/token decode. The h2048/l4 CUDA fp16 run
measured 6.705 ms prefill and 7.333 ms/token decode versus CPU fp32 at
94.036 ms prefill and 30.619 ms/token decode.

\texttt{iree-base-compiler 3.11.0}, \texttt{iree-base-runtime 3.11.0}, and
\texttt{apache-tvm 0.25.0} installed from PyPI. The upstream benchmark command
compiled a StableHLO add module through IREE to a 9,781-byte VMFB and ran
\texttt{iree-benchmark-module}; TVM built and ran a TE vector-add benchmark with
checksum 512.0. A calibrated profile run measured CPU copy at 32.04 GB/s and
CUDA host-to-device transfer at 13.98 GB/s for a 16 MiB sample on this host.

The WSL2 CUDA environment installed NVIDIA CUDA Toolkit 13.3 at
\texttt{/usr/local/cuda}. CacheIR's vLLM launcher now injects the no-UVA fallback
and exports the CUDA toolkit path so FlashInfer JIT can find \texttt{nvcc}. A
real vLLM 0.23.0 latency benchmark with Torch 2.11.0+cu130 and FlashInfer ran
against a CacheIR-created tiny Llama h128 model: 28.535 ms average latency for a
16 input + 8 output token request. The matched CacheIR CPU reference estimate
for the same toy request was 25.425 ms, computed from 10.553 ms prefill and
1.859 ms/token decode. This is a toy batch-1 comparison and does not imply
general superiority over vLLM.

\section{Testing and Validation}

The test suite covers:

\begin{itemize}[leftmargin=*]
  \item decoder graph compilation
  \item decode specialization
  \item pass fusion behavior
  \item artifact JSON round trips
  \item artifact bundle loading
  \item graph export
  \item runtime prefill and decode
  \item CUDA runtime prefill/decode parity with the CPU reference when CUDA is
        available
  \item CUDA scheduler variable-length batched prefill, shared page allocator
        accounting, and active-request batched decode parity with sequential
        CUDA sessions
  \item scheduler queue limits, request priorities, and queued cancellation
  \item prefix-cache-aware prefill equivalence
  \item KV-cache growth
  \item greedy generation
  \item continuous-batch scheduler metrics
  \item server health, metrics, and batch completions
  \item benchmark execution
  \item hardware profile serialization
  \item quantized lowering and runtime compatibility
  \item GGUF dense and \texttt{Q4\_0}/\texttt{Q4\_1}/\texttt{Q5\_0}/\texttt{Q5\_1}/\texttt{Q8\_0}/\texttt{Q8\_1} tensor reads
  \item StableHLO import with reduce-region metadata and MLIR-style dialect export/parser/verifier round trip
  \item calibrated spillover cost model and adapter probe behavior
  \item native pybind11 RMSNorm, matmul, and fused SiLU-multiply parity when the extension is built
\end{itemize}

The latest full local test run passed 34 tests.

Recommended validation commands:

\begin{lstlisting}
python -m pytest -q
python -m compileall cacheir scripts -q
cmake -S cpp -B cpp/build -DCMAKE_BUILD_TYPE=Release
cmake --build cpp/build --config Release
\end{lstlisting}

\section{Complete Technology Stack}

\begin{longtable}{p{0.26\linewidth}p{0.64\linewidth}}
\toprule
\textbf{Area} & \textbf{Technologies} \\
\midrule
Primary languages & Python 3.10+, C++20. \\
Numerical runtime & NumPy reference kernels plus optional PyTorch CUDA tensor
substrate for the CacheIR CUDA executor. \\
Compiler IR & Custom CacheIR graph IR with JSON and textual representations. \\
Model import & Hugging Face config conventions, NPZ, safetensors optional,
ONNX optional, GGUF metadata plus dense \texttt{F32}/\texttt{F16}/\texttt{BF16}/\texttt{I8}/\texttt{I16}/\texttt{I32}/\texttt{I64}/\texttt{F64} and
\texttt{Q4\_0}/\texttt{Q4\_1}/\texttt{Q5\_0}/\texttt{Q5\_1}/\texttt{Q8\_0}/\texttt{Q8\_1} tensor subset, StableHLO textual subset. \\
Transformer semantics & Llama/Mistral/Qwen-style decoder-only blocks,
RMSNorm, RoPE, grouped-query attention, SwiGLU, KV cache. \\
Optimization passes & Shape inference, constant folding, operator fusion,
layout conversion, dead node elimination, mode specialization, quantization-aware
lowering, schedule and memory planning. \\
CPU backend & NumPy reference runtime, C++20/OpenMP native library, scalar,
AVX2/FMA, and AVX512 dispatch, optional pybind11 bridge, native RMSNorm and
fused SiLU-multiply dispatch, opt-in native matmul policy. \\
GPU backend surface & CUDA/Triton target naming, schedule metadata,
\texttt{CudaRuntime}, fp16 CUDA weight loading, GPU-resident KV state, SDPA
attention dispatch, cached packed QKV/Gate-Up weights, shared CUDA page
allocator accounting, variable-length CUDA prefill batching,
scheduler-integrated CUDA decode batching, guarded Triton RMSNorm,
SiLU/SwiGLU, FP16 matmul, fused RMSNorm/QKV/RoPE, single-query decode attention,
and multi-batch page-table decode-attention kernels, optional CUDA C++ fused
kernels, FP16 WMMA Tensor Core matmul, reduced paged-attention decode, batched
paged-attention ABI, and CUDA graph planning. \\
Accelerator adapters & Optional CUTLASS, FlashAttention, and FlashInfer probes,
dispatch contracts, and guarded direct wrappers. \\
Quantization & int4/int8-style lowering and CPU-side quantize/dequantize
simulation. \\
Serving & FastAPI and Uvicorn optional dependencies, OpenAI-compatible
completions/chat, streaming, \texttt{/healthz}, \texttt{/metrics}, CacheIR batch
completions, queue limits, priorities, cancellation, and admission counters. \\
Testing & pytest, compileall, CMake build checks. \\
Benchmarking & CacheIR benchmark CLI with CPU/CUDA backend selector, benchmark
matrix script, CUDA runtime benchmark, CUDA scheduler benchmark, scheduler
benchmark, GPU kernel benchmark, and comparison harness for vLLM, llama.cpp,
IREE, and TVM commands plus installed IREE/TVM smoke benchmark execution and
model-aware vLLM/llama.cpp helpers. \\
Visualization & HTML, DOT, text IR, and MLIR-style CacheIR dialect export plus
parser/verifier round trip. \\
Build and packaging & pyproject.toml, setuptools, CMake, Ninja, Docker,
GitHub Actions. \\
External context & vLLM, llama.cpp, IREE, TVM, StableHLO, MLIR, CUTLASS,
FlashAttention, and FlashInfer are comparison or optional integration surfaces,
not default runtime dependencies. \\
\bottomrule
\end{longtable}

\section{Limitations and Roadmap}

\subsection{Implemented and Tested}

\begin{itemize}[leftmargin=*]
  \item CPU reference execution without PyTorch.
  \item End-to-end CUDA artifact execution through \texttt{CudaRuntime}.
  \item CUDA fp16 weight loading, GPU-resident KV state, SDPA attention dispatch,
        and cached packed QKV/Gate-Up weights.
  \item Scheduler-integrated variable-length CUDA prefill batching and
        active-request CUDA decode batching.
  \item Shared CUDA page allocator accounting across forked request sessions.
  \item Prefill and decode graph specialization.
  \item Paged KV-cache reference behavior.
  \item Prefix-cache reuse with counters and calibrated spillover policy experiments.
  \item Forked per-request runtime sessions sharing weights and tokenizer state.
  \item Continuous-batch scheduler with prefix reuse metrics, queue limits,
        priorities, cancellation, and admission-control counters.
  \item Server health, Prometheus-style metrics, streaming, and batch completions.
  \item Artifact bundles, pass diffs, graph export, benchmarks, and server
        entrypoint.
  \item Quantization-aware lowering with CPU-side quantize/dequantize simulation.
  \item C++20 backend library build with scalar, AVX2/FMA, and AVX512 dispatch.
  \item Optional pybind11 bridge for native RMSNorm, matmul, and fused
        SiLU-multiply kernels.
  \item Native matmul policy with NumPy default and opt-in native force/auto modes.
  \item Guarded Triton RMSNorm, SiLU/SwiGLU, fused RMSNorm/QKV/RoPE,
        single-query decode attention, and multi-batch page-table decode
        attention kernels.
  \item Guarded Triton FP16 matmul using \texttt{tl.dot} Tensor Core lowering
        where available.
  \item Optional CUDA C++ fused kernels for RMSNorm, SwiGLU, RMSNorm/QKV/RoPE,
        FP16 WMMA Tensor Core matmul, batched paged-attention ABI, and reduced
        paged-attention decode.
  \item CUDA graph capture planning for decode replay loops.
  \item Native GGUF dense numeric tensor reads plus classic Q4/Q5/Q8 block
        tensor reads.
  \item Reference GGUF dequantization for upstream-supported K/IQ/TQ/NV/MX formats.
  \item StableHLO text importer with reduce-region metadata.
  \item MLIR-style CacheIR dialect emitter, parser, verifier, and optional
        upstream MLIR C++ dialect registration skeleton.
  \item CUTLASS, FlashAttention, and FlashInfer adapter probes plus guarded
        direct wrappers for prefill, single decode, and batch paged decode.
  \item Direct FlashAttention and FlashInfer smoke execution on WSL2 CUDA
        environments with compatible wheels.
  \item vLLM no-UVA fallback, CacheIR project-root propagation into spawned
        workers, CUDA Toolkit 13.3 path export, and runnable WSL2 CUDA vLLM
        comparison.
  \item Calibrated low-VRAM CPU/GPU spillover cost model with measured bandwidth calibration.
  \item IREE StableHLO compile/runtime benchmark integration through upstream
        IREE wheels.
  \item TVM TE/TIR runtime benchmark integration through the upstream TVM wheel.
  \item Benchmark comparison harness for vLLM, llama.cpp, IREE, and TVM commands,
        including installed IREE/TVM smoke runs and model-aware vLLM/llama.cpp
        helpers.
\end{itemize}

\subsection{Honest Evaluation}

CacheIR is a real, inspectable transformer compiler/runtime project, but it is
not yet a production peer of vLLM, TensorRT-LLM, llama.cpp, or MLC LLM on real
model sizes. The current honest scorecard is:

\begin{longtable}{p{0.42\linewidth}p{0.12\linewidth}p{0.36\linewidth}}
\toprule
\textbf{Dimension} & \textbf{Score} & \textbf{Reason} \\
\midrule
Explainable LLM compiler project & 8/10 & Real IR, passes, artifacts, diffs,
memory plans, imports, tests, and documentation. \\
Research/portfolio systems project & 8/10 & Broad, runnable, benchmarked, and
inspectable. \\
Production LLM serving system & 7/10 & Serving APIs, scheduler, metrics, prefix
reuse, queue limits, priorities, cancellation, CUDA model runtime, shared page
accounting, and variable-length CUDA scheduler batching exist; persistent
page-backed fused attention and reliability hardening are still missing. \\
Raw performance competitor & 6/10 & CUDA fp16 runtime, Triton kernels, batched
decode attention smoke execution, and CUDA scheduler batching show measured
speedups on larger toy shapes, but tuned quantized GEMM and large-model evidence
are still missing. \\
\bottomrule
\end{longtable}

\subsection{Remaining Work Toward an 8/10 Production Score}

\begin{itemize}[leftmargin=*]
  \item Persistent page-backed shared GPU KV storage used directly by scheduler
        fused decode kernels, beyond the current shared allocator accounting.
  \item Fused multi-request paged attention using that persistent shared GPU
        page pool instead of repacking per-request K/V tensors for reference
        kernels.
  \item Packed int4/int8 quantized GEMM with real compression and loader
        integration.
  \item Tuned GEMM paths through CUTLASS, cuBLASLt, Triton autotuning, and
        shape-specific kernel selection.
  \item Production continuous batching hardening: streaming backpressure,
        preemption, fairness policies, bounded latency tests, and long-running
        queue soak tests.
  \item Large-model coverage and benchmarks on 0.5B, 1.5B, and 7B-class local
        models.
  \item Real apples-to-apples throughput and latency comparisons against vLLM,
        llama.cpp, TensorRT-LLM, MLC LLM, IREE, and TVM where those systems are
        installed.
  \item Reliability hardening for memory limits, OOM recovery, tokenizer edge
        cases, model-family conformance, and long-context soak tests.
\end{itemize}

\section{Building PDF or ODF Outputs}

The source of this document is intended to be version controlled. Generated
outputs are intentionally ignored by \texttt{.gitignore}.

To build a PDF with a local LaTeX distribution:

\begin{lstlisting}
pdflatex -interaction=nonstopmode docs/latex/cacheir_project_documentation.tex
\end{lstlisting}

To build an OpenDocument text file with Pandoc:

\begin{lstlisting}
pandoc docs/latex/cacheir_project_documentation.tex \
  -o docs/latex/cacheir_project_documentation.odt
\end{lstlisting}

\section{Conclusion}

CacheIR is a compact but substantial transformer inference compiler project. It
prioritizes clarity, pass-level explainability, prefill/decode specialization,
and KV-cache-aware runtime behavior. The NumPy CPU backend establishes a correct
reference path, while the artifact format, schedules, C++ AVX backend, optional
pybind11 bridge, CUDA artifact runtime, GGUF/StableHLO/MLIR experiments, guarded
Triton kernels, CUDA fused-kernel sources, accelerator adapter probes, and
calibrated spillover model, IREE StableHLO bridge, and TVM benchmark bridge
provide concrete native performance and upstream comparison surfaces. The
project is now strong as an explainable compiler and systems artifact, but
reaching production serving and raw performance parity with mature LLM engines
still requires persistent page-backed shared GPU KV execution, packed quantized
GEMM, tuned large-model CUDA paths, production reliability hardening, and
large-model benchmark evidence.

\end{document}
